\ttitle

### 2015-08-18 Chapters/BodyComp.tex
Body Composition - Major Changes
- Tables changed to include SAT/VAT instead of CSA
- Text changed in most places accordingly
- New tables and figures to reflect new vo2 scalings
- Next commit should finalise this chapter with some luck

\subsection{Body composition and Clinico-pathological characteristics}
Eighty-two patients (35 male) were included in the study. The clinico-pathological characteristics of the study patients and their relationship to body composition is shown in Table \ref{table:bc_clinical} on page \pageref{table:bc_clinical}. There were several significant associations between clinico-pathological variables and body composition as depicted in this table.
	\input{Tables/bc_sex_specific_distrib}
	\input{Tables/bc_clinical}
	\input{Tables/c4_t1}

\subsection{Body Composition in Normal BMI vs Overweight/Obese Patients}
The body composition differences between patients with a normal BMI and patients who are overweight or obese is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}. There were significant differences in the proportion of subcutaneous adipose tissue versus visceral adipose tissue between males and females. Men had generally larger cross-sectional area, less SAT but greater VAT and SM areas. However, the proportion of skeletal muscle in both males and females decreased significantly with increasing BMI.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{Figures/bc_gender_bmi_pie}
	\caption{Differences in body composition according to gender and body mass index.}
	\label{fig:bc_gender_bmi}
\end{figure}

The proportion of skeletal muscle area at L3/L4 decreases from 38\% in male patients with normal BMI to 22\% in males who are obese. There was a greater decrease in the proportion of skeletal muscle area in females with normal BMI (32\%) and obese females (14\%). The higher weight in the high BMI patients was due to a disproportionate increase in adipose tissue rather than skeletal muscle. Moreover, the distribution of the adipose tissue differed between males and females with visceral adipose tissue contributing more to weight in obese males (43\% VAT vs. 35\% SAT) while obese females had a greater proportion of subcutaneous adipose tissue than visceral adipose tissue (68\% SAT vs. 18\% VAT)

@@

Partial correlation analysis was performed to study the relationship between pulmonary function tests and body composition. It has been well-established in previous studies that pulmonary function tests are correlated with age and gender and the analysis was therefore adjusted for these two variables. Forced Vital Capacity (FVC, litres), Forced Expiratory Volume in 1 second (FEV1, litres) and the ratio FEV1/FVC (Tiffeneau-Pinelli index,\%) were compared against the various components of body composition. Both FVC and FEV1 were positively correlated with skeletal muscle area but not with adipose tissue area or total cross-sectional area. FEV1/FVC was not correlated with any of the body composition components. This would indicate that pulmonary function was dependent on skeletal muscle area while FEV1/FVC, a calculated index to quantify restrictive or obstructive lung disease, was not associated with skeletal muscle area. These results are shown in Table \ref{table:bc_cpet} on page \pageref{table:bc_cpet}.

@@

\input{Tables/c4_t2}

Exercise loads achieved at anaerobic threshold and at peak exercise capacity (at volitional stop rather than maximal exercise) were plotted against skeletal muscle area and subcutaneous adipose tissue area measured at L3/L4 to create scatter-plots(Fig. \ref{fig:bc_load_vs_sm}, p\pageref{fig:bc_load_vs_sm}. Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($r^{2} = 0.284, p < 0.001$, Fig. \ref{fig:bc_at_load_vs_sm}) and at peak exercise ($r^{2} = 0.350, p < 0.001$, Fig. \ref{fig:bc_pk_load_vs_sm}). However, no correlation was identified between exercise loads achieved and subcutaneous adipose tissue area either at anaerobic threshold ($r^{2} = 0.004, p = 0.587$) or peak exercise ($r^{2} = 0.020, p = 0.206$).

@@

@@

The correlations between cardiopulmonary exercise parameters and body composition were adjusted for gender. Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak and therefore no adjustments were made for age. The results of this analysis are shown in Table \ref{table:bc_cpet} (p\pageref{table:bc_cpet}).

% Absolute oxygen consumption at the anaerobic threshold in litres per minute was plotted against the total cross-sectional area, subcutaneous adipose tissue area, visceral adipose tissue area and skeletal muscle area. This was repeated by plotting the oxygen consumption at anaerobic threshold corrected for total body weight (in millilitres/kg/min) against the total cross-sectional area, subcutaneous adipose tissue area, visceral adipose tissue area and skeletal muscle area. These scatter-plots are shown in FigureX1. The same was repeated for oxygen consumption at peak exercise and at rest.

Tidal volume ($\dot{V}_T$, litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise. There was a statistically significant but weak positive correlation between minute ventilation ($\dot{V}_E$, litres) and skeletal muscle at anaerobic threshold and peak exercise but not at rest. There was no correlation between either of these measures of pulmonary function and total adipose tissue area at any phase of exercise.

@@

Absolute $\dot{V}_{O_2}$ (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.125, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p<0.001$). However, this correlation was lost after correction of $\dot{V}_{O_2}$ for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.

@@

%$\dot{V}_{O_2}$ corrected to height seems to correlate best with BC and everything else
\clearpage

@@

\subsection{Clinical implications of spurious correlation}
Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}. While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome. However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure. This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}. The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.

>>>>

\subsection{Body composition and Clinico-pathological characteristics}
Eighty-two patients (52 male) were included in the study. 
The clinico-pathological characteristics of the study patients and their relationship to body composition is shown in Table \ref{table:bc_clinical} on page \pageref{table:bc_clinical}. 
There were several significant associations between clinico-pathological variables and body composition as shown in this table. 
Females had a greater area of subcutaneous adipose tissue (223.9 versus 141.6 cm$^2$, p$<$0.001) but less visceral adipose tissue (79.1 versus 174.9 cm$^2$, p$<$0.001).
However, there was no difference in the total adipose tissue area between men and women (316.6 versus 303.0 cm$^2$, p=0.665).
Skeletal muscle area was greater in males than in females (141.3 versus 99.7 cm$^2$).

Patients from more deprived areas of Scotland as defined by the Scottish Index of Multiple Deprivation (SIMD) had a greater amount of subcutaneous adipose tissue (p=0.041) but no difference in the other body composition parameters.
Hypoalbuminemia was associated with less visceral adipose tissue (p=0.013) and less skeletal muscle area that did not reach statistical significance (p=0.054). 
Anaemia (haemoglobin $<$ 12 g/dl) was associated with increased adiposity in the subcutaneous plane and reduced skeletal muscle area. 
	%\input{Tables/bc_sex_specific_distrib}
	%\input{Tables/bc_clinical}
	\input{Tables/c4_t1}

\subsection{Body composition differences between the sexes with increasing BMI}
The body composition differences between male and female patients with increasing body mass index is shown in Figure \ref{fig:bc_gender_bmi} on page \pageref{fig:bc_gender_bmi}. There were significant differences in the proportion of subcutaneous adipose tissue versus visceral adipose tissue between males and females. Men generally had less subcutaneous fat but more visceral fat and skeletal muscle areas. However, the proportion of skeletal muscle in both males and females decreased significantly with increasing body mass index.

\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{Figures/bc_gender_bmi_pie}
	\caption{Differences in body composition according to sex and body mass index.}
	\label{fig:bc_gender_bmi}
\end{figure}

The proportion of skeletal muscle area decreases from 38\% in male patients with normal BMI to 22\% in males who are obese. There was a greater decrease in the proportion of skeletal muscle in females with increasing BMI, with skeletal muscle area decreasing from 32\% in females with normal BMI to 14\% in obese females. The higher weight in the high BMI patients was due to a disproportionate increase in adipose tissue rather than skeletal muscle. Moreover, the distribution of the adipose tissue differed between males and females with visceral adipose tissue contributing more to weight in obese males (43\% VAT vs. 35\% SAT) while obese females had a greater proportion of subcutaneous adipose tissue than visceral adipose tissue (18\% VAT vs. 68\% SAT).

@@

Partial correlation analysis was performed to study the relationship between pulmonary function tests and body composition. It has been well-established in previous studies that pulmonary function tests are correlated with age and gender and the analysis was therefore adjusted for these two variables. Forced Vital Capacity (FVC, litres), Forced Expiratory Volume in 1 second (FEV1, litres) and the ratio FEV1/FVC (Tiffeneau-Pinelli index,\%) were compared against the various components of body composition. Both FVC and FEV1 were positively correlated with skeletal muscle area but not with adipose tissue area or total cross-sectional area. FEV1/FVC was not correlated with any of the body composition components. 
\todo{move this to discussion}
This would indicate that pulmonary function was dependent on skeletal muscle area while FEV1/FVC, a calculated index to quantify restrictive or obstructive lung disease, was not associated with skeletal muscle area. These results are shown in Table \ref{table:bc_cpet} on page \pageref{table:bc_cpet}.

@@

@@

Exercise loads achieved at anaerobic threshold and at peak exercise capacity (at volitional stop rather than maximal exercise) were plotted against skeletal muscle area to create scatter-plots(Fig. \ref{fig:bc_load_vs_sm}, p\pageref{fig:bc_load_vs_sm}. Exercise load correlated positively with skeletal muscle area both at anaerobic threshold ($r^{2} = 0.284, p < 0.001$, Fig. \ref{fig:bc_at_load_vs_sm}) and at peak exercise ($r^{2} = 0.350, p < 0.001$, Fig. \ref{fig:bc_pk_load_vs_sm}). However, no correlation was identified between exercise loads achieved and adipose tissue area either at anaerobic threshold ($r^{2} = 0.004, p = 0.587$) or peak exercise ($r^{2} = 0.020, p = 0.206$) as shown in Table \ref{table:bc_cpet} on p\pageref{table:bc_cpet}.

@@

\input{Tables/bc_cpet}

Partial correlations between cardiopulmonary exercise parameters at rest, anaerobic threshold and peak exercise versus body composition were adjusted for gender. Our own findings (\ref{ch_cpet_jaundice}) and the findings of other authors suggest that age is not related to $\dot{V}_{O_2}$AT or $\dot{V}_{O_2}$Peak and therefore no adjustments were made for age. The results of this analysis are shown in Table \ref{table:bc_cpet} (p\pageref{table:bc_cpet}).

Tidal volume ($\dot{V}_T$, litres) was significantly correlated with skeletal muscle area at all phases of exercise including at rest, anaerobic threshold and peak exercise. There was a statistically significant but weak positive correlation between minute ventilation ($\dot{V}_E$, litres) and skeletal muscle at anaerobic threshold and peak exercise but not at rest. There was no correlation between either of these measures of pulmonary function and adipose tissue area at any phase of exercise.

@@

Absolute $\dot{V}_{O_2}$ (litres/min) had a strong positive correlation with skeletal muscle area at rest ($\rho = 0.353, p = 0.001$), at anaerobic threshold ($\rho = 0.463, p<0.001$) and at peak exercise ($\rho = 0.375, p = 0.001$). However, this correlation was lost after correction of $\dot{V}_{O_2}$ for total body weight and in fact there was a non-significant change in the direction of correlation to the negative.

@@

$O_2$Pulse at the anaerobic thresold and peak exercise was strongly associated with skeletal muscle area with weaker positive correlations with visceral fat and total fat content. There was no correlation between $O_2$Pulse and subcutaneous adipose tissue at any phase of exercise.

%$\dot{V}_{O_2}$ corrected to height seems to correlate best with BC and everything else

\subsection{Scaling $\dot{V}_{O_2}$ to body composition and other factors}

Absolute $\dot{V}_{O_2}$ at the anaerobic threshold was corrected for weight, the square of height, body mass index, skeletal muscle area and estimated lean body mass. Estimated lean body mass (eLBM) was calculated using the Boer formula \parencite{boer_estimated_1984} as shown below.

Men:\ eLBM = 0.407*weight(kg) + 0.267*height(cm) - 19.2

Women:\ eLBM = 0.252*weight(kg) + 0.473*height(cm) - 48.3

The resulting values were compared to body composition, height, weight and body mass index using partial correlations controlling for sex. These results are shown in Table \ref{table:bc_at_new_indices} on p\pageref{table:bc_at_new_indices}. The results from Table \ref{table:bc_cpet} comparing absolute $\dot{V}_{O_2}$ and corrected $\dot{V}_{O_2}$ at the anaerobic threshold with body composition are reproduced in the first two columns of this table to allow comparison with the new indices. $\dot{V}_{O_2}$ scaled for skeletal muscle area had no significant correlation with adipose tissue, height, weight or BMI but had a negative correlation with skeletal muscle area. However, $\dot{V}_{O_2}$ corrected for the square of patient's height had no correlation with adipose tissue but retained a postive correlation with skeletal muscle area. However, there was also a positive correlation with weight and BMI.

However, when absolute $\dot{V}_{O_2}$ was corrected for lean body mass, there was no correlation with any of the body composition indices or with height, weight or body mass index. The negative correlation that occurs between adipose tissue and $\dot{V}_{O_2}$ corrected for total body weight does not occur when $\dot{V}_{O_2}$ is corrected for lean body mass (Fig. \ref{fig:bc_scatter_VO2_TAT_elbm}). Moreover, the correlation with skeletal muscle area remains unchanged regardless of whether $\dot{V}_{O_2}$ is corrected for total body weight or lean body mass (Fig. \ref{fig:bc_scatter_VO2_skeletal_elbm}).

%----------------------------------------------------------------------------------------------
% Correlation between body composition and $\dot{V}_{O_2}$AT before and after correction for total body weight.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_skeletal_elbm}
		\caption{$\dot{V}_{O_2}$AT vs. Skeletal Muscle}
		\label{fig:bc_scatter_VO2_skeletal_elbm}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\centering
		\includegraphics[width=\textwidth]{Figures/bc_scatter_VO2_TAT_elbm}
		\caption{$\dot{V}_{O_2}$AT vs. Total Adipose Tissue}
		\label{fig:bc_scatter_VO2_TAT_elbm}
	\end{subfigure}
	\caption{Correlation between body composition and $\dot{V}_{O_2}$AT corrected for total body weight and estimated lean body mass.}
	\label{fig:bc_scatter_VO2_bodycomp_elbm}
\end{figure}
%----------------------------------------------------------------------------------------------

\input{Tables/bc_at_new_indices}

\clearpage

@@

\subsection{Clinical implications of spurious correlation}

\textit{`Spurious correlation'} is a tern introduced by English mathematician and biometrician, Karl Pearson in 1896. He used the term to describe correlations that occurred as a result of using ratios instead of absolute values rather than because of any actual correlations between the variables being studied\parencite{pearson_mathematical_1896}
Older et al in their pioneering study in 1993 reported that $\dot{V}_{O_2}$AT < 11mls/kg/min was associated with increased mortality in elderly patients undergoing major abdominal surgery \parencite{older_preoperative_1993}. While they did not provide any data on other preoperative or intra-operative factors, they concluded that cardiopulmonary exercise testing was useful in predicting postoperative outcome. However, this first report on the use of cardiopulmonary exercise testing as a preoperative risk assessment tools repeatedly states that a $\dot{V}_{O_2}$AT < 11mls/kg/min represented cardiac failure. This association is repeated in their later work on 548 patients which also showed a clear association between $\dot{V}_{O_2}$AT < 11mls/kg/min and mortality due to cardiovascular causes \parencite{older_cardiopulmonary_1999}. The concepts of \textit{'surgical anaerobic threshold'} and \textit{'postoperative cardiac failure'} were introduced later and were described as the \textit{'inability of the heart to meet the demand of postoperative stress.'}\parencite{society_ats/accp_2003}.
